Open integrated distance sampling for modelling age-structured population dynamics

### 2024-06-26 Manuscript_Nilsen_Nater.qmd

---
title: "An integrated open population distance sampling approach for modelling age-structured populations"
format: docx
author: Erlend B. Nilsen & Chloé R. Nater
editor: visual
bibliography: rypeIDSM_MS1.bib
---

@@

*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution. One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data. Historically, DS models were only available as closed* *population models, and did not allow for the estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates. Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes density. The core of the resulting model is, in essence, equivalent to both an integrated population model and a matrix population model with two age classes: juveniles and adults. The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types. Moreover, as demographic rates often respond to environmental variation, our approach also supports direct estimation of the effects of such environmental covariates on demographic rates. Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics. Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*

@@

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed. Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Since 2019, data has been collected through a mobile app tailormade for this project, which is available trhough App Store (for ios phones) and Google Play (for Android phones). Before 2019, field workers reported their data through a dedicated web portal. After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA). The data are then standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020; @Kvasnes2018; @Nilsen2023a].

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively. Posterior samples from the model fitted to real data are available at @Nilsen2024 (in folder PosteriorSamples_LierneCaseStudy).

@@

Real data, however, are likely to be subject to certain biases, and such biases might result in biased parameter estimates unless accounted for. One type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals. In our case study on willow ptarmigan in Norway, such misclassification is likely to happen at an unknown rate, even if the size difference between adult and juvenile birds are quite substantial during the survey. Moreover, The probability for misclassificiation might be related to both the timing of the survey (e.g. mid August rather than early August), it might vary between observers, and even by survey conditions. Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment. This is due to the way it useses the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see Methods). In the model version presented in the results section, we made the assumption that these birds were in fact juveniles. Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024 ) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival). Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density. Since the proportion of "unknown sex and age" observations in our ptarmigan case study was rather low, potential biases in estimates resulting from age misclassification are expected to be small. Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.

@@

In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$). In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate. Unfortunately, we were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see supplementary figures in @Nilsen2024). This may seem somewhat surprising given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). We speculate that there are at least three potential (non-mutual) explanations to this result. The first is that our covariate data may not have been well suited for estimating effects on recruitment. The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content. While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades [@Kausrud2008; @Cornulier2013], and our study area might be no exception. Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging. Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects. Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that strong unaccounted for effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance. Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates . Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne. @Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.

The new IDSM framework presented here is relevant for many wildlife populations that are surveyed using line transect sampling that includes additional information on age, sex, and/or life stages of the observed individuals. Following the integrated modelling philosophy, the IDSM also allows for the integration of auxiliary data. In our application here, we integrated data from radio-telemetry of marked birds, which explicitly supported the estimation of survival probabilities. The IDSM framework is very flexible, however, and open to the inclusion of additional/other auxiliary data that contains information on demographic rates or population size/density. Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates. Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.

>>>>

---
title: "Open integrated distance sampling for modelling age-structured population dynamics"
format: docx
author: Erlend B. Nilsen & Chloé R. Nater
editor: visual
bibliography: rypeIDSM_MS1.bib
---

@@

*Estimation of abundance and demographic rates for populations of wild species is a challenging but fundamental issue for both management and research into ecology and evolution. One set of approaches that has been used extensively to estimate abundance of wildlife populations is Distance Sampling (DS) methods for line or point transect survey data. Historically, DS models were only available as closed population models, and did not allow for the estimation of changes in abundance through time. The advent of open population formulations based on the DS framework greatly extended the scope of the models, but much untapped potential remains in models that estimate temporal dynamics not only in abundance but also in the underlying demographic rates. Here, we present an integrated distance sampling approach that utilize age-structured survey data and auxiliary data from marked individuals to jointly estimate population density and the demographic rates (recruitment rate and survival probability) that drive temporal changes. The resulting model is equivalent to an integrated population model with a two age classes: juveniles and adults. The integrated framework allows making full use of the available data by effectively combining line transect and telemetry data, and can easily be adapted to include additional and/or different data types. Moreover, as demographic rates often respond to environmental variation, our approach also supports direct estimation of the effects of such environmental covariates on demographic rates. Through a comprehensive simulation study we show that the model is able to adequately recover true population and vital rate dynamics. Subsequent application to data from a study of willow ptarmigan (Lagopus lagopus) in Norway showcases the frameworks ability to recover both fluctuations and trends in population dynamics and highlights its potential applicability to a wide range of species surveyed using distance sampling approaches.*

@@

Line transect survey data were collected in August each year, prior to the annual autumn harvest season, as part of the program "Hønsefuglportalen". Hønsefuglportalen is a national program for line transect surveys of tetraonid birds, and the effort is directed mainly towards willow ptarmigan habitats. In our case study, we used data from the western part of Lierne municipality. Line transects are surveyed by trained volunteers that use pointing dogs to locate the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds in the group, as well as the age (juvenile or adult) and sex of the birds are recorded. As the surveys are conducted in early August, juveniles can be distinguished from adults by their smaller body size. Males and females are manly distinguished by the sound (males often make a characteristic sound when being flushed). Nevertheless, mis-identification can occur, and in addition a proportion of the birds are registered as "unknown" age and/or sex. Besides bird observations, field workers also record whether (1) or not (0) they encounter small rodents on any transect line, allowing the proportion of transect lines with small rodent detection to be used as measure of rodent occupancy (covariate ranging from 0 to 1).Since 2019, data has been collected through a mobile app tailormade for this project, which is available through App Store (for iOS phones) and Google Play (for Android phones). Before 2019, field workers reported their data through a dedicated web portal. After data are collected and reported, they undergo several steps of quality control: first by local contacts and subsequently by personnel at the Norwegian Institute for Nature Research (NINA). The data are then standardized based on the Darwin-Core standard [@Wieczorek2012], and made publicly available as a sampling-event data set published through GBIF [@Nilsen2023a]. For additional description of the data collection procedures, see [@Bowler2020; @Kvasnes2018; @Nilsen2023a].

@@

For the model fits to simulated and real data we ran 3 and 4 MCMC chains with NIMBLE's standard samples for 500k and 100k iterations, respectively. 300k and 40k and thereof were discarded as burn-in prior to thinning with factors 5 and 20 , leaving us with 40k and 3k posterior samples per chain (total of 120k and 12k samples per run), respectively. MCMC parameters were chosen to yield a representative number of samples from converged chains, and convergence was determined based on visual inspection of trace plots. Posterior samples from the model fitted to real data are available at @Nilsen2024 (in folder PosteriorSamples_LierneCaseStudy).

@@

Real data, however, are likely to be subject to certain biases, which may result in biased parameter estimates unless accounted for. One type of bias that is likely to be common for age-structured distance sampling data arises from failure to (correctly) classify the age class of observed individuals. In our case study on willow ptarmigan in Norway, such misclassification is likely to happen at an unknown rate, even if the size difference between adult and juvenile birds are quite substantial during the survey. Moreover, The probability for misclassificiation might be related to both the timing of the survey (e.g. mid August rather than early August), it might vary between observers, and even by survey conditions. Observations with incorrectly classified age have the potential to introduce bias in the IDSMs relative estimates of survival and recruitment. This is due to the way it useses the distance sampling data to estimate survival and recruitment rates. In our process model, the population growth rate ($\lambda$) is determined by the survival and recruitment rate in the following way: $\lambda = S + (S * R)$, and this creates a dependence between the demographic parameters. If the age ratio in the data are biased or contain frequent misclassifications, this is likely to affect the relative contribution of survival and recruitment to the growth rate. To get an idea of the potential effect of this on parameter estimation, we checked the sensitivity of the output of the model fit to real data with regard to the treatment of birds classified as "unknown sex and age" by the field personnel (see Methods). In the model version presented in the results section, we made the assumption that these birds were in fact juveniles. Comparing estimates to an alternative scenario in which we discarded all birds classified as "unknown sex and age" (see supplementary figures in @Nilsen2024 ) we found that -- as expected -- estimated population density was virtually unaffected by the treatment of "unknown sex and age" observations, while the annual demographic parameters shifted proportional to the amount of "unknown sex and age" observations in the given year (towards higher recruitment and lower survival). Thus, biases in the reported age ratios may affect estimated of demographic rates, but not so much population density. Since the proportion of "unknown sex and age" observations in our ptarmigan case study was low (< 3% of observations), potential biases in estimates resulting from age misclassification are expected to be small. Nonetheless, future developments of the IDSM modelling framework should focus on ways of accounting explicitly for misclassification of age class in the field.

@@

Compared to previous studies on ptarmigan (see e.g. [@Israelsen2020; @Sandercock2011]), we could have expected somewhat higher estimates of survival probability. One potential reason for the overall lower estimates obtained here is that our IDSM analysis assumed constant survival over the period 2007-2021 while survival, in reality, must have varied over time. If survival in more recent years, when telemetry data was collected [and the study of e.g. \@ @Israelsen2020 was carried out] was higher than in earlier years -- something that seems likely given population increase over recent years -- an average over the entire time period is expected to be lower. On a different note, we can also not exclude the possibility of a small degree of bias in survival estimates due to misclassification of age in the data (see above), especially seeing as the IDSM's recruitment rate estimates also appear somewhat high compared to other studies on ptarmigan (+refs). <!-- If we have any "biological" thoughts about why somewhat high recruitment rates may not be that ridiculous here, then that could be added here -->

In addition to estimating demographic rates from line transect data, the IDSM also allows including relevant environmental effects on the demographic rates themselves, and not just on population growth rate as a whole ($\lambda$). In the ptarmigan case study we thus attempted to investigate the effect of small rodent abundance (approximated as the proportion of transect lines on which rodents were reported each year) on recruitment rate. We were not able to detect a clear effect of rodent abundance due to large uncertainty associated with the estimate (see Supplementary Figures "Rep_betaR.R.png" in @Nilsen2024). This may seem somewhat surprising given that such a pattern has been reported repeatedly in the literature (see e.g. @Bowler2020). 
We speculate that there are at least three potential and not mutually exclusive explanations to this result. 
The first is that our covariate data may not have been well suited for estimating effects on recruitment. The data on rodent abundance was heavily zero-inflated, and the annual variation in the index was rather small otherwise, making for a covariate with relatively little information content. While this may be partially a consequence of how these data are collected, it is also well known that the amplitude and regularity of the rodent cycles has been fading in recent decades [@Kausrud2008; @Cornulier2013], and our study area might be no exception. 
Lack of peak rodent years in the time series to which we fitted the model may thus also have contributed to making effect estimation challenging. Second, it is possible that rodent effects were obscured by other, potentially stronger, covariate effects. Previous research has shown that ptarmigan recruitment is also sensitive to the weather in the late winter and spring (before and during the breeding season); as we did not fit any weather covariates to the model so there is a possibility that effects of spring conditions in certain years may have masked any remaining effects of small rodent abundance. Finally, the data set used in this analysis is relatively short (15 years), leaving us with somewhat limited statistical power to detect effects of temporal covariates . Taken together we therefore do not consider this study as a particularly strong test of the underlying effect of small rodent fluctuations on ptarmigan recruitment rates. It is worth noting that future applications could increase statistical power by including either more years of data or capitalizing on space-for-time substitution as the Norwegian ptarmigan monitoring programme spans many more locations beyond Lierne. @Bowler2020, for example, used data from the same sampling program but from more areas using a simpler open population DS model, and detected a very clear signal from small rodent abundance on ptarmigan population growth rate. An extension of our IDSM to include data from multiple areas therefore constitutes a promising approach for investigating to which extent similar results emerge by linking environmental covariates to the actual demographic rates and not only just to the resulting population growth rates.

The new IDSM framework presented here is relevant for many wildlife populations that are surveyed using line transect sampling that includes additional information on age, sex, and/or life stages of the observed individuals. Following the integrated modelling philosophy, the IDSM also allows for the integration of auxiliary data. In our application here, we integrated data from radio-telemetry of marked birds, which explicitly supported the estimation of survival probabilities. The IDSM framework is very flexible, however, and open to the inclusion of additional/other auxiliary data that contains information on demographic rates or population size/density. Moreover, hierarchical nature of the model makes is straight forward to adapt to different species and to include different suites of environmental covariates on the demographic rates. Finally, it constitutes a modelling framework that is well suited for extension to multiple areas and thus able to capitalize on space-for-time substitution [@lovell2023] to produce large-scale and spatially explicit estimates of population density, demographic rates, and environmental effects from large-scale (participatory) monitoring.
